Fast Fourier transformation of electromagnetic data for computationally expensive kernels

### 2021-03-20 ManuscriptTime.tex
references
WOI

\rtodo 3 Reply; Include comments from earlier review; papers that were
referred; difficulties in late/early times; IP effects. Refer to steel-casing
effects (lindsey), but state we look here at non-Debye cole models.
Add also the papers seogi mentioned re IP effects.
\emph{We tried to improve the introduction and focus more on the actual topic.
  Amongst other, we address the topic of steel casing and non-Debye Cole-Cole
models.}

@@

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains.
Another approach is to have Maxwell's equations only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the
frequency domain to the time domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies \added{or times}, translating into
a significant reduction in computation time. The required computation grids for
low frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required\replaced{, which}{. We build our approach upon
\mbox{\cite{GEO.07.Plessix}}, who presented such an adaptive gridding for
multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain} is naturally based on the skin depth,
\deleted{which is} the distance after which the amplitude of the
electromagnetic field is decayed by $1/e\approx 37\,\%$. The skin depth
$\delta$ is a function of conductivity and frequency, and for the diffusive
approximation of Maxwell's equations in an isotropic, homogeneous medium is
given by \citep[e.g.,][ equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\added{We build our approach upon} \cite{GEO.07.Plessix}\added{, who presented
such an adaptive gridding for multi-frequency (and multi-source) CSEM
modelling. They} define the minimum cell width $\Delta_\mathrm{min}$ as
a fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this yields quite large cells for low frequencies, so special care has
to be taken around the source by defining a maximum allowed
$\Delta_\mathrm{min}$. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the buffer zone outside of the
area where source and receivers are located.

@@

\added{The requirement for any transform is a robust solver of sufficient
precision.} \replaced{We use f}{F}or the numerical computations \deleted{we
use} the open-source (Apache License 2.0) code\added{s \empymod
\mbox{\citep{GEO.17.Werthmuller}} and} \emg3d
\citep{JOSS.19.Werthmuller}\replaced{.}{,} \added{The former computes
semi-analytical responses of layered models, the latter is} a multigrid solver
based on \cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov
subspace solvers or as a solver on its own. The multigrid approach works fine
for the diffusive approximation of Maxwell's equations, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All the examples in this article are run on a laptop
with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using Ubuntu
20.04 and Python 3.8. All examples were run using a single thread, but
parallelization over frequencies is straight forward, as the computations are
independent. \deleted{For the semi-analytical computations of layered models we
use the open-source (Apache License 2.0) code \empymod
\mbox{\citep{GEO.17.Werthmuller}}.}

@@

\added{mention independence of spatial complexity, why we also show 1D
examples.}

>>>>

\rtodo 3 Reply; Include comments from earlier review; papers that were
referred; difficulties in late/early times; IP effects. Refer to steel-casing
effects (lindsey), but state we look here at non-Debye cole models.
Add also the papers seogi mentioned re IP effects. \cite{CAG.19.Heagy}
\cite{EXP.20.Kang} \emph{We tried to improve the introduction and focus more on
  the actual topic. Amongst other, we address the topic of steel casing and
  non-Debye Cole-Cole models.}

@@

\rtodo \added{REDO and send to R.R}

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains.
Another approach is to have Maxwell's equations only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the
frequency domain to the time domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies\replaced{ through a combination of
extrapolation, interpolation, and setting the responses for high frequencies to
zero. This translates}{, translating} into a significant reduction in
computation time. \replaced{The number of required frequencies is the most important aspect for reducing computation time. The second most important aspect is the computation grid, as t}{T}he required \deleted{computation} grids
for low frequencies (in our case in the order of 0.001\,Hz) and high
frequencies (around 100\,Hz) are hugely different. Low frequencies can be
computed on a coarser grid, but they require a much larger domain in order to
avoid boundary effects. High frequencies, on the other hand, require denser
gridding, but they are much more limited in reach. An adaptive gridding scheme
is therefore required\replaced{, which}{. We build our approach upon
\mbox{\cite{GEO.07.Plessix}}, who presented such an adaptive gridding for
multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain} is naturally based on the skin depth,
\deleted{which is} the distance after which the amplitude of the
electromagnetic field is decayed by $1/e\approx 37\,\%$. The skin depth
$\delta$ is a function of conductivity and frequency, and for the diffusive
approximation of Maxwell's equations in an isotropic, homogeneous medium is
given by \citep[e.g.,][\mbox{\deleted{ equation~1.53}}]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\added{We build our approach upon} \cite{GEO.07.Plessix}\added{, who presented
such an adaptive gridding for multi-frequency (and multi-source) CSEM
modelling. They} define the minimum cell width $\Delta_\mathrm{min}$ as
a fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this \replaced{can yield}{yields} quite large cells for low
frequencies, so special care has to be taken around the source by defining a
maximum allowed $\Delta_\mathrm{min}$. The grid dimension, on the other hand,
is defined as a function of \deleted{the} skin depth for the average
conductivity of the background, $\delta_\mathrm{ave}$. They use four times
$\delta_\mathrm{ave}$ for the $x$-, $y$-, and downward $z$-directions, and a
fixed 50\,km for the upward $z$-direction to account for the airwave. To reduce
the number of cells, it is desirable to introduce stretching, at least in the
buffer zone outside of the area where source and receivers are located.

@@

\added{The requirement for any transform is a robust solver of sufficient
precision over a wide range of frequencies.} \replaced{We use f}{F}or the
numerical computations \deleted{we use} the open-source (Apache License 2.0)
code\added{s \empymod} \added{\mbox{\citep{GEO.17.Werthmuller}}} \added{and}
\emg3d \citep{JOSS.19.Werthmuller}\replaced{.}{,} \added{The former computes
semi-analytical responses of layered models, the latter is} a \added{3D}
multigrid solver based on \cite{GP.06.Mulder}, which can be used as a
preconditioner for Krylov subspace solvers or as a solver on its own. The
multigrid approach works fine for the diffusive approximation of Maxwell's
equations, which assumes that $\omega\varepsilon \ll \sigma$, where
$\varepsilon$ is electric permittivity (F/m). The remaining system to solve in
the frequency domain is then given by the second-order differential equation of
the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All \deleted{the} examples in this article are run on a
laptop with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using
Ubuntu 20.04 and Python 3.8. \replaced{They}{All examples} were run using a
single thread, but parallelization over frequencies is straight forward, as the
computations are independent. \deleted{For the semi-analytical computations of
layered models we use the open-source (Apache License 2.0) code \empymod
\mbox{\citep{GEO.17.Werthmuller}}.}

@@

\added{mention independence of spatial complexity, why we also show 1D
examples; however, sufficiently accurate 3D results are an obvious
pre-requisite.}
